### 2017-11-13 Draft2/Appending.Dynamic.tex
Remove [ref1]
Algorithms
Styling for the algos
Work towards other sections
Replace Var with Lcl for algo local variables
Algos: Variables for param names
Algos: Add spaces between commas
Work

I will implement appending for dynamic arrays first. Let $\VarList$ be a dynamic array. The following definitions are used in the code:

\begin{description}
	\item[initial capacity] Denoted by $\VarInitCapacity$. The capacity of an empty dynamic array.\\
	{\HdrAssumptions} $\VarInitCapacity$ is an integer, $\VarInitCapacity > 0$
	\item[growth factor] Denoted by $\VarGrowthFactor$. The factor by which the current capacity is multiplied to get the new capacity when $\VarList$ is non-empty and grows.\\
	{\HdrAssumptions} $\VarGrowthFactor\VarInitCapacity \geq \VarInitCapacity + 1$
\end{description}

\begin{algorithm}
	\begin{algorithmic}[1]
		\Procedure{$\FuncAppend$}{$L, item$}
			\If{$\VarList.\FieldFull$}
				\State $\VarList.\FuncGrow()$
			\EndIf

			\State $\VarList.\FieldHead[\VarList.\FieldHeadSize] \gets item$
			\State $\VarList.\FieldSize \gets \VarList.\FieldSize + 1$
		\EndProcedure
		\Statex
		\Procedure{$\FuncGrow$}{$L$}
			\State $new\ buf \gets \FuncNewArray(\VarList.\FieldSize \times \VarGrowthFactor)$
			\State $\FuncArrayCopy(\VarList.\FieldBuffer, new\ buf, \VarList.\FieldSize)$
			\State $\VarList.\FieldBuffer \gets new\ buf$
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

@@

Before I analyze the time complexity of $\FuncAppend$, I consider a different method for measuring its cost. Suppose I start with an empty collection and $n$ elements are appended. How many times is an element stored in an array? I will term the answer to this question the \textbf{write cost} of $n$ appends, and denote it $\FuncWrites(n)$.

@@

	The final value of the sequence, $\VarCapacitySeq_\VarUseful$, is the capacity needed for $n$ items. By assumption (3), $\VarCapacitySeq_\VarUseful \geq n > \VarCapacitySeq_{\VarUseful - 1}$. Consider the case when $n > \VarInitCapacity$: it must be true that $\VarCapacitySeq_\VarUseful > \VarInitCapacity$, so $\VarUseful \geq 1$. Since $\VarUseful - 1 \neq 0$, $\VarCapacitySeq_\VarUseful = \VarGrowthFactor^\VarUseful\VarInitCapacity$ and $\VarCapacitySeq_{\VarUseful - 1} = \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity$. Then
	\begin{align*}
	\VarGrowthFactor^\VarUseful\VarInitCapacity &\geq n > \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
	\VarGrowthFactor^\VarUseful &\geq \frac{n}{\VarInitCapacity} > \VarGrowthFactor^{\VarUseful - 1}\\
	\VarUseful &\geq \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity > \VarUseful - 1\\
	\end{align*}
	Since $\VarUseful$ is an integer,
	\begin{align*}
	\VarUseful &= \ExprMessy
	\end{align*}
	Now consider the case when $n \leq \VarInitCapacity$. By assumption (3), $n > \VarCapacitySeq_{\VarUseful - 1}$. $\VarUseful - 1$ must then equal $-1$, since any other value would imply $n > \VarCapacitySeq_{\VarUseful - 1} \geq \VarInitCapacity$. Thus $\VarUseful = 0$.

	It was shown $\VarUseful \geq 1 \geq 0$ for the first case, and it can be shown $\ExprMessy \leq 0$ for the second case. Then, a general formula for $\VarUseful$ is as follows:
	\begin{align*}
	\VarUseful &= \max(\ExprMessy, 0)
	\end{align*}
	The final term in the sequence is $\VarGrowthFactor^\VarUseful\VarInitCapacity = \VarGrowthFactor^{\max(\ExprMessy, 0)}\VarInitCapacity$, completing the proof.
\end{proof}

\begin{corollary}
\label{coro:GrowthSeq}
	Let the \textbf{growth sequence}, $\VarGrowSeq$, of $\VarList$ be the sizes for which $\FuncGrow$ is called when $n$ items are appended. Then $\VarGrowSeq = \VarCapacitySeq \setminus \left\{ \VarCapacitySeq_\VarUseful \right\}$.
\end{corollary}

\begin{proof}
	If $\VarCapacitySeq_i$ exists and $i \geq 1$, then clearly $\FuncGrow$ must have been called when the size was $\VarCapacitySeq_{i - 1}$, so $\VarCapacitySeq_{i - 1} \in \VarGrowSeq$. Then $\VarGrowSeq$ contains every term in $\VarCapacitySeq$ except for the last, $\VarCapacitySeq_\VarUseful$, as the corollary states.
\end{proof}

When $\FuncGrow$ is called and the current size is $\VarGrowSeq_i$, the algorithm copies $\VarGrowSeq_i$ items. Then the total number of items copied when $n$ items are appended is:
\begin{align*}
\sum_i {\VarGrowSeq_i} &= \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \ldots + \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Counting the writes made for each item by $\FuncAppend$, an explicit formula for $\FuncWrites(n)$ is as follows:
[ref1]
\begin{align*}
\FuncWrites(n) = n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Now, my goal is to approximate $\FuncWrites(n)$ with $\sim$. To make is easier to do so, I will asymptotically bound $\VarGrowthFactor^\VarUseful$ which depends on $n$.

\begin{lemma}
\label{lem:ToVarUsefulPowerBounds}
	For $n > \VarInitCapacity$, $\dfrac n \VarInitCapacity \leq \VarGrowthFactor^\VarUseful < \dfrac {\VarGrowthFactor{n}} \VarInitCapacity$.
\end{lemma}

\begin{proof}
	It was shown in Lemma \ref{lem:CapacitySeq} that if $n > \VarInitCapacity$, $\VarUseful = \ExprMessy \geq 1$.  Now note that $\VarUseful$ may also be written as $\ExprMessyAlt$. Then
	\begin{align*}
	\ExprMessyAltInner \leq \VarUseful &< \ExprMessyAltInner + 1\\
	\frac{n}{\VarInitCapacity} \leq \VarUseful &< \frac{\VarGrowthFactor{n}}{\VarInitCapacity}
	\end{align*}
	as desired.
\end{proof}

Now, I proceed to asymptotically bound $\FuncWrites(n)$.
\begin{align*}
\FuncWrites(n) &= n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{n / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \FluteLeq \FuncWrites(n) &\FluteLess n + \left( \frac{\VarGrowthFactor{n} / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n - \left( \frac{\VarInitCapacity}{\VarGrowthFactor - 1} \right) \FluteLeq \FuncWrites(n) &\FluteLess \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n - \left( \frac{\VarInitCapacity}{\VarGrowthFactor - 1} \right)\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n \FluteLeq \FuncWrites(n) &\FluteLess \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n
\end{align*}
\HdrSpaceComplex

>>>>

I will implement appending for dynamic arrays first. Let $\VarList$ be a dynamic array. The following constants are used by the algorithm:

\begin{algorithm}
	\begin{algorithmic}[1]
		\Procedure{$\FuncAppend$}{$\VarList,\ \ParamItem$}
			\If{$\VarList.\FieldFull$}
				\State $\VarList.\FuncGrow()$
			\EndIf

			\State $\VarList.\FieldBuffer[\VarList.\FieldSize] \gets \ParamItem$
			\State $\VarList.\FieldSize \gets \VarList.\FieldSize + 1$
		\EndProcedure
		\Statex
		\Procedure{$\FuncGrow$}{$\VarList$}
			\State $\LclNewBuffer \gets \FuncNewArray(\VarGrowthFactor \times \VarList.\FieldSize)$
			\State $\FuncArrayCopy(\VarList.\FieldBuffer,\ \LclNewBuffer,\ \VarList.\FieldSize)$
			\State $\VarList.\FieldBuffer \gets \LclNewBuffer$
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

@@

% Todo: CITE
It is well-known that appending one item to a dynamic array takes amortized $O(1)$ time. Thus, appending $n$ items takes $O(n)$ time. However, this fact is not very useful for the purpose of comparison--- later, it will be shown that growth arrays also take $O(n)$ time to append $n$ items.

It would be helpful if the ratio $\dfrac{\FuncTimeDynamicArray(n)}{\FuncTimeGrowthArray(n)}$ could be determined, where $\FuncTimeDynamicArray(n)$ and $\FuncTimeGrowthArray(n)$ denote the average time needed to append $n$ items for dynamic and growth arrays, respectively. However, there is no way to quantify these functions without empirical measurements. Thus, I will approximate this ratio through a different technique: Suppose $n$ elements are appended to an empty collection. How many times is an element stored in an array? I will term the answer to this question the \textbf{write cost} of $n$ appends, and denote it $\FuncWrites(n)$.

I will na\"{i}vely assume that $\FuncTime(n) \propto \FuncWrites(n)$, or that the average time for $n$ appends is proportional to their write cost. Then $\dfrac{\FuncTimeDynamicArray(n)}{\FuncTimeGrowthArray(n)} = \dfrac{\FuncWritesDynamicArray(n)}{\FuncWritesGrowthArray(n)}$, where $\FuncWritesDynamicArray$ and $\FuncWritesGrowthArray$ are respectively the write cost functions for dynamic and growth arrays. In this section, I will derive the formula for $\FuncWritesDynamicArray(n)$. Because this section concerns dynamic arrays only, I will write it as $\FuncWrites(n)$.

@@

	The final value of the sequence, $\VarCapacitySeq_\VarUseful$, is the capacity needed to store $n$ items. By assumption (3), $\VarCapacitySeq_\VarUseful \geq n > \VarCapacitySeq_{\VarUseful - 1}$. Consider the case when $n > \VarInitCapacity$: it must be true that $\VarCapacitySeq_\VarUseful > \VarInitCapacity$, so $\VarUseful \geq 1$. Since $\VarUseful - 1 \geq 0$, $\VarCapacitySeq_\VarUseful = \VarGrowthFactor^\VarUseful\VarInitCapacity$ and $\VarCapacitySeq_{\VarUseful - 1} = \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity$. Then
	\begin{align*}
	\VarGrowthFactor^\VarUseful\VarInitCapacity &\geq n > \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
	\VarGrowthFactor^\VarUseful &\geq \frac{n}{\VarInitCapacity} > \VarGrowthFactor^{\VarUseful - 1}\\
	\VarUseful &\geq \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity > \VarUseful - 1\\
	\end{align*}
	Since $\VarUseful$ is an integer,
	\begin{align*}
	\VarUseful &= \ExprMessy
	\end{align*}
	Now, consider the case when $n \leq \VarInitCapacity$. By assumption (3), $n > \VarCapacitySeq_{\VarUseful - 1}$. $\VarUseful - 1$ must then equal $-1$, since any other value would imply $n > \VarCapacitySeq_{\VarUseful - 1} \geq \VarInitCapacity$. Thus $\VarUseful = 0$.

	It was shown that $\VarUseful \geq 1 \geq 0$ for the first case, and it can be shown that $\ExprMessy \leq 0$ for the second case. Therefore, a general formula for $\VarUseful$ is as follows:
	\begin{align*}
	\VarUseful &= \max(\ExprMessy, 0)
	\end{align*}
	The final term in the sequence is $\VarGrowthFactor^\VarUseful\VarInitCapacity = \VarGrowthFactor^{\max(\ExprMessy, 0)}\VarInitCapacity$, completing the proof.
\end{proof}

\begin{corollary}
\label{coro:GrowthSeq}
	Let the \textbf{growth sequence}, $\VarGrowSeq$, of $\VarList$ be the sizes at which $\FuncGrow$ is called when $n$ items are appended to $\VarList$. Then $\VarGrowSeq = \VarCapacitySeq \setminus \left\{ \VarCapacitySeq_\VarUseful \right\}$.
\end{corollary}

\begin{proof}
	% Todo: Better language here? 'exists' is confusing
	If $\VarCapacitySeq_i$ exists and $i \geq 1$, then clearly $\FuncGrow$ must have been called when the size was $\VarCapacitySeq_{i - 1}$, so $\VarCapacitySeq_{i - 1} \in \VarGrowSeq$. Then $\VarGrowSeq$ contains every term in $\VarCapacitySeq$ except for the last, $\VarCapacitySeq_\VarUseful$, as the corollary states.
\end{proof}

When $\FuncGrow$ is called and the current size is $\VarGrowSeq_i$, the algorithm copies $\VarGrowSeq_i$ items. Then the total number of items copied when $n$ items are appended is:
\begin{align*}
\sum_i {\VarGrowSeq_i} &= \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \ldots + \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Counting the writes made per item by $\FuncAppend$, an explicit formula for $\FuncWrites(n)$ is as follows:
\begin{align*}
\FuncWrites(n) = n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Now, my goal is to approximate $\FuncWrites(n)$ with $\sim$. To make this easier to do, I will asymptotically bound $\VarGrowthFactor^\VarUseful$ with respect to $n$.

\begin{lemma}
\label{lem:ToVarUsefulPowerBounds}
	$\dfrac n \VarInitCapacity \FluteLeq \VarGrowthFactor^\VarUseful \FluteLess \dfrac {\VarGrowthFactor{n}} \VarInitCapacity$. That is, $\dfrac n \VarInitCapacity \leq \VarGrowthFactor^\VarUseful < \dfrac {\VarGrowthFactor{n}} \VarInitCapacity$ for sufficiently large $n$.
\end{lemma}

\begin{proof}
	It was shown in Lemma \ref{lem:CapacitySeq} that if $n > \VarInitCapacity$, $\VarUseful = \ExprMessy \geq 1$.  Now, note that $\VarUseful$ may be written as $\ExprMessyAlt$ for such $n$. Then,
	\begin{align*}
	\ExprMessyAltInner \leq \VarUseful &< \ExprMessyAltInner + 1\\
	\frac{n}{\VarInitCapacity} \leq \VarUseful &< \frac{\VarGrowthFactor{n}}{\VarInitCapacity}
	\end{align*}
	as desired.
\end{proof}

I now proceed to asymptotically bound $\FuncWrites(n)$:
\begin{align*}
\FuncWrites(n) &= n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{n / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \FluteLeq \FuncWrites(n) &\FluteLess n + \left( \frac{\VarGrowthFactor{n} / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n - \left( \frac{\VarInitCapacity}{\VarGrowthFactor - 1} \right) \FluteLeq \FuncWrites(n) &\FluteLess \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n - \left( \frac{\VarInitCapacity}{\VarGrowthFactor - 1} \right)\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n \FluteLeq \FuncWrites(n) &\FluteLess \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n
\end{align*}
\HdrSpaceComplex

### 2017-11-13 Draft2/AsymptoticNotation.tex
Work

I will also introduce two cousins of $\sim$, $\FluteLess$ and $\FluteLeq$. $f \FluteLeq g$ if and only if $f \FluteLess g$ or $f \sim g$. $\FluteLess$ is defined as follows:
\begin{align*}
f \FluteLess g \Biconditional \ExprNToInfty \frac f g < 1
\end{align*}
When reading these symbols, one says ``$f$ is asymptotically less than $g$'' or ``$f$ is asymptotically less than or equal to $g$''.

@@

Here, I define various properties of the $\sim$ relation that later proofs will use. The properties themselves are proved in the appendix.

@@

\begin{corollary}
\label{coro:BothSides}
	Both sides of a $\sim$ relation may be added, multiplied, or divided by the same function.
\end{corollary}

@@

The following theorem states that if two functions are asymptotic, then they can be used interchangeably in an asymptotic inequality.

@@

\begin{corollary}
\label{coro:InterchangeableInNonStrictInequality}
	The same statements hold true for $\FluteLeq$: If $f \sim \Tweak{f}$ and $\Tweak{f} \FluteLeq \TweakSecond{f}$, then $f \FluteLeq \TweakSecond{f}$. If $g \sim \TweakSecond{g}$ and $\Tweak{g} \FluteLeq \TweakSecond{g}$, then $\Tweak{g} \FluteLeq g$.
\end{corollary}

% Todo: Add note, adding doesn't always work.
In the same vein as Corollary \ref{coro:BothSides}, the following theorem states that asymptotic inequalities can be algebraically manipulated.

\begin{theorem}
\label{thm:BothSidesInequality}
	Both sides of a $\FluteLess$ or $\FluteLeq$ relation may be multiplied or divided by the same function.
\end{theorem}

>>>>

I will also introduce two cousins of $\sim$, $\FluteLess$ and $\FluteLeq$. They are defined as follows:
\begin{align*}
f \FluteLess g &\Biconditional \ExprNToInfty \frac f g < 1\\
f \FluteLeq g &\Biconditional f \FluteLess g \lor f \sim g
\end{align*}
I will call these relations \textbf{asymptotic inequalities}. The above inequalities are read ``$f$ is asymptotically less than $g$'' and ``$f$ is asymptotically less than or equal to $g$,'' respectively.

@@

I define various properties of the $\sim$ relation here, which later proofs will use. The properties themselves are proved in the appendix.

@@

\begin{corollary}
\label{coro:BothSides}
	A $\sim$ equation holds if both sides are added, multiplied, or divided by the same quantity.
\end{corollary}

@@

The following theorem states that if two functions are asymptotic, then they can be interchanged in an asymptotic inequality.

@@

\begin{corollary}
\label{coro:InterchangeableInNonStrictInequality}
	This theorem also holds for $\FluteLeq$: If $f \sim \Tweak{f}$ and $\Tweak{f} \FluteLeq \TweakSecond{f}$, then $f \FluteLeq \TweakSecond{f}$. If $g \sim \TweakSecond{g}$ and $\Tweak{g} \FluteLeq \TweakSecond{g}$, then $\Tweak{g} \FluteLeq g$.
\end{corollary}

The following theorem states that asymptotic inequalities can be algebraically manipulated, in the same vein as Corollary \ref{coro:BothSides}. However, they do not always hold if the same quantity is added to both sides; for example, $1 \FluteLess 2$, but $n + 1 \sim n + 2$.

\begin{theorem}
\label{thm:BothSidesInequality}
	A $\FluteLess$ or $\FluteLeq$ inequality holds if both sides are multiplied or divided by the same quantity.
\end{theorem}

### 2017-11-13 Draft2/CommonOperations.Blurb.tex
Work

In this section, I implement common operations for dynamic and growth arrays, and analyze their time complexity. I also analyze space complexity of operations that allocate memory.

>>>>

In this section, I implement commonly used operations for dynamic and growth arrays, then analyze their time complexity. I also analyze the space complexity of operations that allocate memory.

### 2017-11-13 Draft2/FieldsProperties.tex
Work

\begin{itemize}
	\item $\VarList.\FieldBuffer$ - The \textbf{buffer}, or raw array, that $\VarList$ stores its items in.
	\item $\VarList.\FieldSize$ - The number of items in $\VarList$.
\end{itemize}

@@

\begin{itemize}
	\item $\VarList.\FieldHead$ - The \textbf{head} of $\VarList$. It returns the buffer new items are appended to.
	\item $\VarList.\FieldTail$ - The \textbf{tail} of $\VarList$. (\textbf{Important note: The tail is a dynamic array.}) It returns a dynamic array of pointers to buffers that are already filled with items. The tail can be thought of as a two-dimensional array.\\
	({\HdrRemark} It may seem strange for a growth array to use the very data structure it is replacing. Lemma \ref{lem:VarUsefulIsOLogN} shows, however, that only $O(\log n)$ many pointers are appended to the tail. Thus, the extra copying and allocations the tail performs is minuscule compared to other work done by the growth array.)
	\item $\VarList.\FieldSize$ - The number of items in $\VarList$.
	\item $\VarList.\FieldCapacity$ - The \textbf{capacity} of $\VarList$. It returns the maximum number of items $\VarList$ can hold before it must grow.
\end{itemize}

>>>>

\begin{itemize}
	\item $\VarList.\FieldBuffer$ -- The \textbf{buffer}, or raw array, that $\VarList$ stores its items in.
	\item $\VarList.\FieldSize$ -- The number of items in $\VarList$.
\end{itemize}

@@

\begin{itemize}
	\item $\VarList.\FieldHead$ -- The \textbf{head} of $\VarList$. It returns the buffer new items are appended to.
	\item $\VarList.\FieldTail$ -- The \textbf{tail} of $\VarList$. (\textbf{Important note: The tail is a dynamic array.}) It returns a dynamic array of pointers to buffers that are already filled with items. The tail can be thought of as a two-dimensional array.\\
	({\HdrRemark} It may seem strange for a growth array to use the very data structure it is replacing. Lemma \ref{lem:VarUsefulIsOLogN} will show, however, that only $O(\log n)$ many pointers are appended to the tail. Thus, the extra copying and allocations the tail performs is minuscule compared to other work done by the growth array.)
	\item $\VarList.\FieldSize$ -- The number of items in $\VarList$.
	\item $\VarList.\FieldCapacity$ -- The \textbf{capacity} of $\VarList$. It returns the maximum number of items $\VarList$ can hold before it must grow.
\end{itemize}
